---
title: "BIP Summer-Intro to R Workshop Activity"
author: "Joscelin Rocha-Hidalgo"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE, warning=FALSE)
```

I am super excited to walk you through our activity report while I show you how you can use some functions and even create some basic visualizations!

## How did I choose the questions?

You may be wondering why I asked you the questions you saw in the online survey. I am a developmental scientist and most of my current research is with children between the ages of 4 and 6 years. Some of them helped me come up with some of the "funner" questions such as the ones asking about superpowers. And of course, I also chose some questions that could give me some variability of responses and one that is very poorly stated so we can do some data cleaning on its content.

## Let's load some packages

For the purposes of this workshop, most of our functions will come from packages part of the bigger package called *"Tidyverse"*.

```{r}
#Load the `tidyverse` package. 
library(tidyverse)

#Load the `janitor` package.
library(janitor)
```

## Let's import our dataset

In theory, we could have imported the googlesheets directly to R (definitely possible!) but for simplicity and lack of time, I downloaded the responses as a csv and saved them to our repo under the "data" folder.

```{r }
#Import the data.csv file into R and assign it to an object called "data"
#FYI: you could have called it anything!
data <- read_csv("data/data.csv")

#After you run it, check the environment panel. It should be saved there!
```
## Let's clean & rename the variables
The variables exported from the googlesheet are too long and we need to `rename()` them to shorter ones.
```{r}
#let's clean the variables names using the package `janitor()

data <- data %>%
  janitor::clean_names()

## You can run this code to get the column names so it is easier to copy and paste 
## them when changing them
#colnames(data)  


data <- data %>%
  rename( "books_2023" = "how_many_books_have_you_read_so_far_this_2023",
          "enjoy_cooking" = "on_a_scale_of_1_to_10_how_much_do_you_enjoy_cooking",
          "hrs_sleep" = "how_many_hours_of_sleep_do_you_get_on_average_per_night",
          "superpower" = "which_superpower_would_you_choose",
          "chosen_weekend_activity" = "if_traveling_was_not_an_issue_you_can_teleport_anywhere_in_a_blink_what_would_be_your_preferred_way_to_spend_a_free_weekend",
          "least_fav_movie" = "whats_your_least_favorite_type_of_movie",
          "age_group" = "what_is_your_age_group",
          "gender" = "what_is_your_gender",
          "state" = "which_state_do_you_currently_reside_in",
          "race" = "what_is_your_racial_background",
          "latin_status" = "are_you_latino_a_e_x"  )
```


## Let's separate the demo questions and the "fun" questions into 2 new dataframes

To separate the demo questions and the fun questions into 2 separate dfs, we will use the function `select()`.

1.  Assign the fun questions to a df called "fun_data"

2.  Assign the demo question to a df called "demo_data"
```{r}
# you can list them one by one
data %>%
  select(age_group,race)

#another way:
data %>%
  select(age_group:latin_status)

#another way
data %>%
  select(age_group:last_col())

```

